---
title: "BIOS629 Final Project: Clinical vs. Genomic Predictors of Overall Survival in NSCLC"
author: "Wenjie Han"
date: "2026-04-18"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 10
    fig_height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center'
)

# Load required libraries
library(tidyverse)
library(survival)
library(survminer)
library(randomForestSRC)
library(survcomp)
library(caret)
library(gridExtra)

# Set seed for reproducibility
set.seed(629)
```

# 1. Introduction

## 1.1 Research Question

This study investigates whether genomic features can improve prediction of overall survival outcomes in non-small cell lung cancer (NSCLC) beyond standard clinical variables. We extend our interim analysis by implementing **5-fold cross-validation** to obtain more reliable performance estimates and **risk stratification** to enhance clinical interpretability.

## 1.2 Data Source

- **Dataset**: MSK-CHORD (Memorial Sloan Kettering Cancer Center, Nature 2024)
- **Population**: Patients with NSCLC
- **Sample Size**: N = 7,809

# 2. Data Loading and Preprocessing

## 2.1 Read Data

```{r load-data}
# MSK-CHORD dataset (Nature, 2024)
# Source: https://www.cbioportal.org/study/summary?id=msk_chord_2024
# The NSCLC subset is read directly from this repository's GitHub raw URL.
data <- read.csv(
  "https://raw.githubusercontent.com/Havenhan02/BIOS629-Final-Project/main/dataset/msk_chord_2024_NSCLC_data.csv",
  skip = 1
)

cat("Dataset dimensions:", dim(data)[1], "rows x", dim(data)[2], "columns\n")
```

## 2.2 Select and Clean Variables

```{r select-variables}
# Select relevant variables for analysis
analysis_data <- data %>%
  select(
    age = `Current.Age`,
    sex = Sex,
    stage = `Stage..Highest.Recorded.`,
    fga = `Fraction.Genome.Altered`,
    smoking = `Smoking.History..NLP.`,
    mutation_count = `Mutation.Count`,
    os_time = `Overall.Survival..Months.`,
    os_status = `Overall.Survival.Status`,
    tmb = `TMB..nonsynonymous.`,
    tumor_purity = `Tumor.Purity`
  )

# Display initial summary
cat("Initial data summary:\n")
summary(analysis_data)
```

## 2.3 Data Cleaning and Transformation

```{r data-cleaning}
# Clean and transform variables
analysis_data <- analysis_data %>%
  mutate(
    # Create binary event indicator (1 = deceased, 0 = censored)
    event = ifelse(grepl("DECEASED|1:DECEASED", os_status), 1, 0),
    
    # Convert categorical variables to factors
    stage = as.factor(stage),
    sex = as.factor(sex),
    
    # Ensure numeric variables are properly formatted
    age = as.numeric(age),
    fga = as.numeric(fga),
    mutation_count = as.numeric(mutation_count),
    os_time = as.numeric(os_time),
    tmb = as.numeric(tmb),
    tumor_purity = as.numeric(tumor_purity)
  ) %>%
  # Remove rows with missing survival time
  filter(!is.na(os_time) & os_time > 0)

# Set factor levels for smoking history
analysis_data$smoking <- factor(
  analysis_data$smoking,
  levels = c("Never", "Former/Current Smoker", "Unknown")
)

# Display cleaned data summary
cat("\nCleaned data summary:\n")
cat("Final sample size:", nrow(analysis_data), "\n")
cat("Number of events (deaths):", sum(analysis_data$event), "\n")
cat("Censoring rate:", round(100 * (1 - mean(analysis_data$event)), 1), "%\n")
```

## 2.4 Handle Missing Data

```{r missing-data}
# Check missing data pattern
missing_summary <- analysis_data %>%
  summarise(
    across(everything(), ~sum(is.na(.)))
  ) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  mutate(Missing_Percent = round(100 * Missing_Count / nrow(analysis_data), 2)) %>%
  filter(Missing_Count > 0) %>%
  arrange(desc(Missing_Count))

print(missing_summary)

# For this analysis, we use complete cases for the core variables
# Note: tumor_purity has substantial missing data, so we'll focus on other variables
analysis_complete <- analysis_data %>%
  select(-tumor_purity) %>%  # Remove due to high missingness
  filter(
    !is.na(age), 
    !is.na(sex), 
    !is.na(stage), 
    !is.na(fga), 
    !is.na(smoking),
    !is.na(mutation_count),
    !is.na(tmb)
  )

cat("\nFinal complete case sample size:", nrow(analysis_complete), "\n")
```

# 3. Exploratory Data Analysis

## 3.1 Descriptive Statistics

```{r descriptive-stats}
# Continuous variables
continuous_vars <- analysis_complete %>%
  select(age, fga, mutation_count, tmb, os_time)

cat("Continuous Variables Summary:\n")
summary(continuous_vars)

# Categorical variables
cat("\n\nCategorical Variables:\n")
cat("\nSex distribution:\n")
table(analysis_complete$sex)

cat("\nStage distribution:\n")
table(analysis_complete$stage)

cat("\nSmoking history distribution:\n")
table(analysis_complete$smoking)

cat("\nEvent distribution:\n")
table(analysis_complete$event)
```

## 3.2 Kaplan-Meier Survival Analysis

### Overall Survival Curve

```{r km-overall, fig.width=10, fig.height=7}
km_fit <- survfit(Surv(os_time, event) ~ 1, data = analysis_complete)

km_overall_plot <- ggsurvplot(
  km_fit,
  data = analysis_complete,
  risk.table = TRUE,
  conf.int = TRUE,
  title = "Overall Survival Curve for NSCLC Patients",
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  surv.median.line = "hv",
  palette = "#2E86AB",
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE,
  ncensor.plot = TRUE,
  ggtheme = theme_minimal(base_size = 12)
)

print(km_overall_plot)
```

### Survival by Stage

```{r km-stage, fig.width=12, fig.height=8}
km_stage <- survfit(Surv(os_time, event) ~ stage, data = analysis_complete)

km_stage_plot <- ggsurvplot(
  km_stage,
  data = analysis_complete,
  pval = TRUE,
  pval.method = TRUE,
  risk.table = TRUE,
  title = "Survival by Cancer Stage",
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  palette = "nejm",
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE,
  ncensor.plot = TRUE,
  ggtheme = theme_minimal(base_size = 12),
  legend.title = "Stage",
  legend.labs = levels(analysis_complete$stage)
)

print(km_stage_plot)
```

### Survival by Sex

```{r km-sex, fig.width=10, fig.height=7}
km_sex <- survfit(Surv(os_time, event) ~ sex, data = analysis_complete)

km_sex_plot <- ggsurvplot(
  km_sex,
  data = analysis_complete,
  pval = TRUE,
  risk.table = TRUE,
  title = "Survival by Sex",
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  palette = c("#E64B35", "#4DBBD5"),
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE,
  ggtheme = theme_minimal(base_size = 12)
)

print(km_sex_plot)
```

# 4. Model Development and Cross-Validation

## 4.1 Create 5-Fold Cross-Validation Folds

```{r create-folds}
# Create 5-fold cross-validation folds
folds <- createFolds(analysis_complete$event, k = 5, list = TRUE)

cat("Cross-validation fold sizes:\n")
for(i in 1:5) {
  cat("Fold", i, ":", length(folds[[i]]), "observations\n")
}
```

## 4.2 Define Models

We will compare three models:
1. **Clinical Cox Model**: age + sex + stage + smoking
2. **Clinical + Genomic Cox Model**: age + sex + stage + smoking + fga + mutation_count + tmb
3. **Random Survival Forest**: All variables with non-linear relationships

## 4.3 Cross-Validation Performance Evaluation

```{r cv-function}
# Function to calculate C-index for a Cox model on test data
calculate_cox_cindex <- function(model, test_data) {
  # Get linear predictor (risk scores)
  lp <- predict(model, newdata = test_data, type = "lp")
  
  # Calculate concordance
  concord <- survConcordance(Surv(test_data$os_time, test_data$event) ~ lp)
  return(concord$concordance)
}

# Function to calculate C-index for RSF on test data
calculate_rsf_cindex <- function(model, test_data) {
  # Predict on test data
  pred <- predict(model, newdata = test_data)
  
  # Calculate concordance using predicted mortality
  concord <- survConcordance(Surv(test_data$os_time, test_data$event) ~ pred$predicted)
  return(concord$concordance)
}
```

```{r run-cv, results='hide', message=FALSE, warning=FALSE}
# Initialize results storage
cv_results <- data.frame(
  Fold = integer(),
  Model = character(),
  C_index = numeric(),
  stringsAsFactors = FALSE
)

# Store predictions for risk stratification
all_predictions <- data.frame(
  row_index = integer(),
  fold = integer(),
  cox_clinical_lp = numeric(),
  cox_full_lp = numeric(),
  rsf_risk = numeric(),
  os_time = numeric(),
  event = numeric()
)

# Run 5-fold cross-validation
for(i in 1:5) {
  cat("Processing Fold", i, "...\n")
  
  # Split data
  test_idx <- folds[[i]]
  train_data <- analysis_complete[-test_idx, ]
  test_data <- analysis_complete[test_idx, ]
  
  # Model 1: Clinical Cox
  cox_clinical <- coxph(
    Surv(os_time, event) ~ age + sex + stage + smoking,
    data = train_data,
    x = TRUE, y = TRUE
  )
  cox_clinical_cindex <- calculate_cox_cindex(cox_clinical, test_data)
  cox_clinical_lp <- predict(cox_clinical, newdata = test_data, type = "lp")
  
  # Model 2: Clinical + Genomic Cox
  cox_full <- coxph(
    Surv(os_time, event) ~ age + sex + stage + smoking + fga + mutation_count + tmb,
    data = train_data,
    x = TRUE, y = TRUE
  )
  cox_full_cindex <- calculate_cox_cindex(cox_full, test_data)
  cox_full_lp <- predict(cox_full, newdata = test_data, type = "lp")
  
  # Model 3: Random Survival Forest
  rsf_model <- rfsrc(
    Surv(os_time, event) ~ age + sex + stage + smoking + fga + mutation_count + tmb,
    data = train_data,
    ntree = 500,
    importance = FALSE,
    forest = TRUE
  )
  rsf_cindex <- calculate_rsf_cindex(rsf_model, test_data)
  rsf_pred <- predict(rsf_model, newdata = test_data)
  rsf_risk <- rsf_pred$predicted  # Predicted mortality
  
  # Store results
  cv_results <- rbind(cv_results, data.frame(
    Fold = i,
    Model = "Clinical Cox",
    C_index = cox_clinical_cindex
  ))
  
  cv_results <- rbind(cv_results, data.frame(
    Fold = i,
    Model = "Clinical + Genomic Cox",
    C_index = cox_full_cindex
  ))
  
  cv_results <- rbind(cv_results, data.frame(
    Fold = i,
    Model = "Random Survival Forest",
    C_index = rsf_cindex
  ))
  
  # Store predictions for risk stratification
  fold_predictions <- data.frame(
    row_index = test_idx,
    fold = i,
    cox_clinical_lp = cox_clinical_lp,
    cox_full_lp = cox_full_lp,
    rsf_risk = rsf_risk,
    os_time = test_data$os_time,
    event = test_data$event
  )
  
  all_predictions <- rbind(all_predictions, fold_predictions)
}

cat("Cross-validation completed!\n")
```

## 4.4 Cross-Validation Results Summary

```{r cv-summary}
# Summarize CV results
cv_summary <- cv_results %>%
  group_by(Model) %>%
  summarise(
    Mean_C_index = mean(C_index),
    SD_C_index = sd(C_index),
    Min_C_index = min(C_index),
    Max_C_index = max(C_index),
    .groups = 'drop'
  ) %>%
  arrange(desc(Mean_C_index))

print(cv_summary)

# Display fold-by-fold results
print(cv_results %>% pivot_wider(names_from = Model, values_from = C_index))
```

## 4.5 Visualize Cross-Validation Results

```{r cv-plot, fig.width=12, fig.height=7}
# Box plot of C-index across folds
cv_plot <- ggplot(cv_results, aes(x = Model, y = C_index, fill = Model)) +
  geom_boxplot(alpha = 0.7, width = 0.5) +
  geom_jitter(width = 0.1, size = 3, alpha = 0.7) +
  scale_fill_manual(values = c("#4DBBD5", "#00A087", "#E64B35")) +
  coord_cartesian(ylim = c(0.60, 0.75)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "5-Fold Cross-Validation: Model Performance Comparison",
    subtitle = "C-index distribution across folds",
    y = "C-index",
    x = NULL
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 15, hjust = 1)
  )

print(cv_plot)
```

```{r cv-bar-plot, fig.width=10, fig.height=6}
# Bar plot of mean C-index with error bars
cv_bar_plot <- ggplot(cv_summary, aes(x = reorder(Model, Mean_C_index), y = Mean_C_index, fill = Model)) +
  geom_bar(stat = "identity", width = 0.6, alpha = 0.8) +
  geom_errorbar(aes(ymin = Mean_C_index - SD_C_index, ymax = Mean_C_index + SD_C_index), 
                width = 0.2, linewidth = 1) +
  scale_fill_manual(values = c("#4DBBD5", "#00A087", "#E64B35")) +
  coord_cartesian(ylim = c(0.60, 0.75)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Mean C-index from 5-Fold Cross-Validation",
    subtitle = "Error bars represent standard deviation across folds",
    y = "C-index",
    x = NULL
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  geom_text(aes(label = sprintf("%.3f", Mean_C_index)), vjust = -1, size = 5, fontface = "bold")

print(cv_bar_plot)
```

# 5. Final Model Training and Risk Stratification

## 5.1 Train Final Models on Full Dataset

```{r final-models}
# Train final models on complete dataset for interpretation
cat("Training final models on full dataset...\n")

# Final Clinical Cox Model
final_cox_clinical <- coxph(
  Surv(os_time, event) ~ age + sex + stage + smoking,
  data = analysis_complete,
  x = TRUE, y = TRUE
)

cat("\n=== Clinical Cox Model Summary ===\n")
summary(final_cox_clinical)

# Final Clinical + Genomic Cox Model
final_cox_full <- coxph(
  Surv(os_time, event) ~ age + sex + stage + smoking + fga + mutation_count + tmb,
  data = analysis_complete,
  x = TRUE, y = TRUE
)

cat("\n=== Clinical + Genomic Cox Model Summary ===\n")
summary(final_cox_full)

# Final Random Survival Forest Model
final_rsf <- rfsrc(
  Surv(os_time, event) ~ age + sex + stage + smoking + fga + mutation_count + tmb,
  data = analysis_complete,
  ntree = 1000,
  importance = TRUE,
  forest = TRUE
)

cat("\n=== Random Survival Forest Summary ===\n")
print(final_rsf)
```

## 5.2 Variable Importance Analysis

```{r variable-importance, fig.width=12, fig.height=8}
# Calculate variable importance from RSF
vimp_result <- vimp(final_rsf)

# Create importance data frame
importance_df <- data.frame(
  Variable = names(vimp_result$importance),
  Importance = as.numeric(vimp_result$importance)
) %>%
  arrange(desc(Importance))

print(importance_df)

# Plot variable importance
importance_plot <- ggplot(importance_df, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_bar(stat = "identity", fill = "#2E86AB", alpha = 0.8) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  labs(
    title = "Variable Importance from Random Survival Forest",
    subtitle = "Higher values indicate greater importance for prediction",
    x = "Variable",
    y = "Importance Score"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(importance_plot)
```

## 5.3 Generate Risk Scores and Stratify Patients

```{r risk-stratification}
# Merge predictions back to original data
analysis_complete$row_index <- 1:nrow(analysis_complete)
analysis_risk <- analysis_complete %>%
  left_join(all_predictions %>% select(row_index, cox_clinical_lp, cox_full_lp, rsf_risk), 
            by = "row_index")

# For patients not in CV predictions (shouldn't happen with proper CV), 
# use final model predictions
missing_idx <- which(is.na(analysis_risk$rsf_risk))
if(length(missing_idx) > 0) {
  analysis_risk$cox_clinical_lp[missing_idx] <- predict(final_cox_clinical, type = "lp")[missing_idx]
  analysis_risk$cox_full_lp[missing_idx] <- predict(final_cox_full, type = "lp")[missing_idx]
  analysis_risk$rsf_risk[missing_idx] <- final_rsf$predicted.oob[missing_idx]
}

# Create risk groups based on tertiles
analysis_risk <- analysis_risk %>%
  mutate(
    # Risk groups based on RSF predictions (tertiles)
    rsf_risk_group = cut(rsf_risk, 
                         breaks = quantile(rsf_risk, probs = c(0, 1/3, 2/3, 1)),
                         labels = c("Low Risk", "Intermediate Risk", "High Risk"),
                         include.lowest = TRUE),
    
    # Risk groups based on Clinical Cox predictions
    cox_clinical_risk_group = cut(cox_clinical_lp, 
                                  breaks = quantile(cox_clinical_lp, probs = c(0, 1/3, 2/3, 1)),
                                  labels = c("Low Risk", "Intermediate Risk", "High Risk"),
                                  include.lowest = TRUE),
    
    # Risk groups based on Full Cox predictions
    cox_full_risk_group = cut(cox_full_lp, 
                              breaks = quantile(cox_full_lp, probs = c(0, 1/3, 2/3, 1)),
                              labels = c("Low Risk", "Intermediate Risk", "High Risk"),
                              include.lowest = TRUE)
  )

# Display risk group distributions
cat("RSF Risk Group Distribution:\n")
print(table(analysis_risk$rsf_risk_group))

cat("\nClinical Cox Risk Group Distribution:\n")
print(table(analysis_risk$cox_clinical_risk_group))

cat("\nFull Cox Risk Group Distribution:\n")
print(table(analysis_risk$cox_full_risk_group))
```

## 5.4 Kaplan-Meier Analysis by Risk Groups

### RSF Risk Groups

```{r km-rsf-risk, fig.width=12, fig.height=8}
km_rsf <- survfit(Surv(os_time, event) ~ rsf_risk_group, data = analysis_risk)

km_rsf_plot <- ggsurvplot(
  km_rsf,
  data = analysis_risk,
  pval = TRUE,
  pval.method = TRUE,
  risk.table = TRUE,
  title = "Survival by RSF Risk Groups",
  subtitle = "Patients stratified by Random Survival Forest predicted risk",
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  palette = c("#00A087", "#E69F00", "#E64B35"),
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE,
  ncensor.plot = TRUE,
  ggtheme = theme_minimal(base_size = 12),
  legend.title = "Risk Group",
  legend.labs = c("Low Risk", "Intermediate Risk", "High Risk")
)

print(km_rsf_plot)
```

### Clinical Cox Risk Groups

```{r km-cox-clinical-risk, fig.width=12, fig.height=8}
km_cox_clinical <- survfit(Surv(os_time, event) ~ cox_clinical_risk_group, data = analysis_risk)

km_cox_clinical_plot <- ggsurvplot(
  km_cox_clinical,
  data = analysis_risk,
  pval = TRUE,
  pval.method = TRUE,
  risk.table = TRUE,
  title = "Survival by Clinical Cox Risk Groups",
  subtitle = "Patients stratified by clinical variables only",
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  palette = c("#00A087", "#E69F00", "#E64B35"),
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE,
  ncensor.plot = TRUE,
  ggtheme = theme_minimal(base_size = 12),
  legend.title = "Risk Group",
  legend.labs = c("Low Risk", "Intermediate Risk", "High Risk")
)

print(km_cox_clinical_plot)
```

### Full Cox Risk Groups

```{r km-cox-full-risk, fig.width=12, fig.height=8}
km_cox_full <- survfit(Surv(os_time, event) ~ cox_full_risk_group, data = analysis_risk)

km_cox_full_plot <- ggsurvplot(
  km_cox_full,
  data = analysis_risk,
  pval = TRUE,
  pval.method = TRUE,
  risk.table = TRUE,
  title = "Survival by Clinical + Genomic Cox Risk Groups",
  subtitle = "Patients stratified by clinical and genomic variables",
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  palette = c("#00A087", "#E69F00", "#E64B35"),
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE,
  ncensor.plot = TRUE,
  ggtheme = theme_minimal(base_size = 12),
  legend.title = "Risk Group",
  legend.labs = c("Low Risk", "Intermediate Risk", "High Risk")
)

print(km_cox_full_plot)
```

# 6. Model Comparison Summary

## 6.1 Performance Metrics Summary

```{r performance-summary}
# Create comprehensive performance summary
performance_summary <- data.frame(
  Model = c("Clinical Cox", "Clinical + Genomic Cox", "Random Survival Forest"),
  CV_Mean_C_index = cv_summary$Mean_C_index[match(c("Clinical Cox", "Clinical + Genomic Cox", "Random Survival Forest"), cv_summary$Model)],
  CV_SD_C_index = cv_summary$SD_C_index[match(c("Clinical Cox", "Clinical + Genomic Cox", "Random Survival Forest"), cv_summary$Model)],
  Full_Data_C_index = c(
    as.numeric(concordance(final_cox_clinical)$concordance),
    as.numeric(concordance(final_cox_full)$concordance),
    1 - final_rsf$err.rate[length(final_rsf$err.rate)]
  )
)

print(performance_summary)

# Calculate improvement over clinical model
clinical_cindex <- performance_summary$CV_Mean_C_index[1]
performance_summary$Improvement_Over_Clinical <- performance_summary$CV_Mean_C_index - clinical_cindex
performance_summary$Relative_Improvement <- round(100 * performance_summary$Improvement_Over_Clinical / clinical_cindex, 2)

print(performance_summary)
```

## 6.2 Final Comparison Plot

```{r final-comparison-plot, fig.width=12, fig.height=7}
# Create comparison plot
comparison_data <- performance_summary %>%
  select(Model, CV_Mean_C_index, CV_SD_C_index) %>%
  rename(C_index = CV_Mean_C_index, SD = CV_SD_C_index)

final_plot <- ggplot(comparison_data, aes(x = reorder(Model, C_index), y = C_index, fill = Model)) +
  geom_bar(stat = "identity", width = 0.6, alpha = 0.8) +
  geom_errorbar(aes(ymin = C_index - SD, ymax = C_index + SD), 
                width = 0.2, linewidth = 1) +
  scale_fill_manual(values = c("#4DBBD5", "#00A087", "#E64B35")) +
  coord_cartesian(ylim = c(0.60, 0.75)) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Final Model Performance Comparison",
    subtitle = "5-Fold Cross-Validation Mean C-index with Standard Deviation",
    y = "C-index",
    x = NULL,
    caption = "Higher C-index indicates better predictive performance"
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    axis.text.x = element_text(angle = 15, hjust = 1, size = 11)
  ) +
  geom_text(aes(label = sprintf("%.3f", C_index)), vjust = -1.5, size = 5, fontface = "bold")

print(final_plot)
```

# 7. Discussion and Conclusions

## 7.1 Key Findings

```{r key-findings}
cat("=== KEY FINDINGS ===\n\n")

cat("1. MODEL PERFORMANCE:\n")
for(i in 1:nrow(performance_summary)) {
  cat(sprintf("   - %s: C-index = %.3f (SD = %.3f)\n", 
              performance_summary$Model[i], 
              performance_summary$CV_Mean_C_index[i],
              performance_summary$CV_SD_C_index[i]))
}

cat("\n2. IMPROVEMENT OVER CLINICAL MODEL:\n")
cat(sprintf("   - Clinical + Genomic Cox: +%.3f (%.1f%% improvement)\n",
            performance_summary$Improvement_Over_Clinical[2],
            performance_summary$Relative_Improvement[2]))
cat(sprintf("   - Random Survival Forest: +%.3f (%.1f%% improvement)\n",
            performance_summary$Improvement_Over_Clinical[3],
            performance_summary$Relative_Improvement[3]))

cat("\n3. RISK STRATIFICATION:\n")
# Calculate median survival by risk group
risk_survival <- analysis_risk %>%
  group_by(rsf_risk_group) %>%
  summarise(
    n = n(),
    median_os = median(os_time),
    events = sum(event),
    event_rate = round(100 * sum(event) / n(), 1),
    .groups = 'drop'
  )

for(i in 1:nrow(risk_survival)) {
  cat(sprintf("   - %s: n=%d, median OS=%.1f months, event rate=%.1f%%\n",
              risk_survival$rsf_risk_group[i],
              risk_survival$n[i],
              risk_survival$median_os[i],
              risk_survival$event_rate[i]))
}

cat("\n4. MOST IMPORTANT PREDICTORS (from RSF):\n")
for(i in 1:min(5, nrow(importance_df))) {
  cat(sprintf("   %d. %s: %.4f\n", i, importance_df$Variable[i], importance_df$Importance[i]))
}
```

## 7.2 Conclusions

Based on our comprehensive analysis using 5-fold cross-validation:

1. **Genomic features improve prediction**: Adding genomic variables (FGA, mutation count, TMB) to clinical variables provides modest but consistent improvement in predictive performance.

2. **Random Survival Forest performs best**: The RSF model achieved the highest C-index (mean = `r round(performance_summary$CV_Mean_C_index[3], 3)`), demonstrating the value of capturing non-linear relationships and variable interactions.

3. **Risk stratification is effective**: All three models successfully stratified patients into distinct risk groups with significantly different survival outcomes (all p < 0.001).

4. **Stage remains the strongest predictor**: Cancer stage is consistently the most important variable across all models, followed by genomic features.

## 7.3 Limitations and Future Work

**Limitations:**
- Cross-validation was performed on the same dataset; external validation on independent cohorts is needed
- Some genomic variables had missing data (e.g., tumor purity)
- The analysis focused on overall survival; progression-free survival analysis could provide additional insights

**Future Directions:**
- External validation using independent NSCLC datasets
- Incorporation of additional genomic biomarkers (e.g., specific driver mutations)
- Exploration of deep learning survival models
- Development of a clinical risk calculator tool

# 8. Session Information

```{r session-info}
sessionInfo()
```
